---
title: "ESR Reference Range Study"
output: html_notebook
---

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
```

## Background and Objectives

The question to be addressed is whether the reference ranges for ESR values obtained at our hospital are appropriate. There is great variation in ESR reference ranges in use by different clinical labs.

The goal of this analysis is to review our hospital's ESR values to find out whether there are discernable normal and abnormal populations. The objective is to visualize a sample of ESR values, broken down by age and sex, and highlight different options for cutoffs to better understand whether adjusting our reference range might be beneficial.

#### Our hospital

|Age|Male|Female|
|---|----|------|
|<=50|<=15 mm/h|<=25 mm/h|
|>50|<=15 mm/h|<=30 mm/h|

#### ARUP Laboratories

|Age|Male|Female|
|---|----|------|
|all|<=10 mm/h|<=20 mm/h|

#### Quest Diagnostics

|Age|Male|Female|
|---|----|------|
|<=50|<=15 mm/h|<=20 mm/h|
|>50|<=20 mm/h|<=30 mm/h|

#### LabCorp

|Age|Male|Female|
|---|----|------|
|<=50|<=15 mm/h|<=32 mm/h|
|>50|<=30 mm/h|<=40 mm/h|

#### Bakerman's ABCs of Interpretive Laboratory Data

|Age|Male|Female|
|---|----|------|
|Newborn|<=2 mm/h|<= 2 mm/h|
|Neonates and children|3 - 13 mm/h|3 - 13 mm/h|
|<40|1 - 15 mm/h|1 - 20 mm/h|
|>=40|Age / 2|(Age / 2) + 5|

## Data Acquisition

Data source: CERNER query performed by Dan Herman on 08/14/2018, then de-identified by the following actions:

1. Replacing patient names and MRNs with synthetic data generated using [Synthea](https://github.com/synthetichealth/synthea)
2. Scrmabling order ID and accession ID
3. Randomly shifting all time stamps by +/- 3 weeks

```{r, message=FALSE}
esr <- read_csv("esr.csv")
```

We will focus on ESR values from 2017, along with the following columns:

1. **MRN**. Patient MRN (**synthetic**).
2. **PT_NAME**. Patient name (**synthetic**).
3. **AGE**. Patient age at time of collection.
4. **SEX**. Patient sex.
6. **RESULT**. Result value. Renamed from **ASCII_TEXT** for clarity.
7. **DRAWN_DT_TM**. Date and time of collection (**shifted randomly**).

```{r}
esr <- esr %>%
  select(MRN,
         PT_NAME,
         AGE,
         SEX,
         DRAWN_DT_TM,
         RESULT = ASCII_TEXT) %>%
  filter(year(DRAWN_DT_TM) == 2017) %>%
  arrange(DRAWN_DT_TM)
```


## Data Exploration and Cleaning

### Missing values

Count the number of `NA`s (missing values) in each column of `esr`.

```{r}
esr %>%
  map_df(function(x) sum(is.na(x))) %>%
  gather(column, NAs)
```

Only the `RESULT` column has missing values. We will explore this later.


### MRN

`MRN`, the HUP MRNs of the patients, is a **character** column, which is appropriate. 

*Note*: if `MRN` were converted to integer, MRNs with leading zeroes would be altered. 

```{r}
set.seed(1)
esr %>%
  sample_n(10) %>%
  pull(MRN)
```

A random sample of 10 MRNs shows that all are 9 digits long. This is what's expected for our patients.


### PT_NAME

`PT_NAME` is a **character** column, which is appropriate. 

```{r}
set.seed(1)
esr %>%
  sample_n(10) %>%
  pull(PT_NAME)
```

A random sample of 10 names is as expected. (Again, these are all fake).


### AGE

`AGE`, the age at collection, is an **integer** column, which is appropriate.

```{r}
ggplot(data = esr) + 
  geom_histogram(mapping = aes(x = AGE),
                 binwidth = 1)
```

The distribution of `AGE` shows a mode at around 70 years, with a left shoulder. A possible explanation is that we see a cohort of younger patients who are worked up for inflammatory conditions and a larger cohort of older patients who are worked up for joint pain. 


### SEX

`SEX` is a **character** column, but since this is a categorial variable, we will convert it to a **factor**.

```{r}
esr <- esr %>%
  mutate(SEX = as_factor(SEX))
  
esr %>%
  group_by(SEX) %>%
  count()
```

The majority (>60%) of ESR tests were performed on female (`F`) patients. 


### DRAWN_DT_TM

`DRAWN_DT_TM` is a **POSIXct** (datetime) column, which is appropriate.

```{r}
ggplot(data = esr) +
  geom_histogram(mapping = aes(x = DRAWN_DT_TM),
                 bins = 365)
```

Weekly cycling of ESR test volume is apparent. Weekly test volume appears approximately constant throughout 2017, without clear peaks or troughs.


### RESULT

Before doing anything else, let's review rows where `RESULT` has missing values.

```{r}
esr %>%
  filter(is.na(RESULT))
```

No clear pattern emerges. It might be interesting to follow up to find out the reason for missingness. However, since only 41/36723 (0.1%) of the data are affected, it seems reasonable to drop these rows:

```{r}
esr <- esr %>% 
  drop_na(RESULT)
```


`RESULT` is a **character** column but should be an **integer**, so we will convert it. 

Before doing so, examine any rows that do *not* contain an integer value:

```{r, warning=FALSE}
esr %>%
  filter(RESULT %>% as.integer() %>% is.na) %>%
  group_by(RESULT) %>%
  summarize(n = n()) %>% 
  arrange(desc(n))
```

About 1700 rows (out of about 37,000) have a non-integer value in `RESULT` - about **4%**. The large majority of these have a value of `< 1`. Dropping these results would skew the distribution to the right. It's better to replace each value `< 1` with `0`. 

```{r}
esr <- esr %>%
  mutate(RESULT = case_when(
    RESULT == "< 1" ~ "0",
    TRUE ~ RESULT
  ))
```

Now convert `RESULT` to integer and drop all remaining rows with a non-integer `RESULT` value. 

```{r, warning=FALSE}
esr <- esr %>%
  mutate(RESULT = as.integer(RESULT)) %>%
  drop_na(RESULT)
```

```{r}
ggplot(data = esr) + 
  geom_histogram(mapping = aes(RESULT))
```

Apparently we are dealing with some outliers. 

```{r}
esr %>%
  filter(RESULT > 150) %>%
  arrange(desc(RESULT))
```

Very few high values, possibly transcription errors.

Remove values >150:

```{r}
esr <- esr %>%
  filter(RESULT <= 150)
```

```{r}
ggplot(data = esr) + 
    geom_histogram(mapping = aes(RESULT),
                   binwidth = 1)
```

The distribution of ESR values appears to have a mode at around 10, with a long right tail, which represents the abnormal values. 

There are conspicuous peaks at regular intervals (multiples of 5), suggestive of the tendency of humans to record "nice" numbers.


### Overview of the Cleaned Data

```{r, warning=FALSE}
esr
```


## Visualization and Modeling

Do normal and abnormal ESR values naturally separate into distinct populations, when broken down by age and sex?

```{r}
ggplot(data = esr) +
  geom_point(mapping = aes(x = AGE, y = RESULT),
             size = 0.5, 
             alpha = 0.1) + 
  facet_wrap(~SEX) + 
  labs(x = "Patient Age (years)",
       y = "ESR (mm/h)")
```

There is no clear separation of normal and abnormal values. In addition, there are more data points on the `Female` panel, which makes this graph difficult to interpret. 


Will a separation become apparent when equal numbers of points are shown on each graph, and a density contour plot is overlaid?

```{r}
set.seed(1)
esr_sample <- esr %>%
  group_by(SEX) %>%
  sample_n(13000)

ggplot(data = esr_sample, 
       mapping = aes(x = AGE, 
                     y = RESULT)) +
  facet_wrap(~SEX) + 
  geom_point(size = 0.5, 
             alpha = 0.1) +
  geom_density2d(color = "deepskyblue") +
  labs(x = "Patient Age (years)",
       y = "ESR (mm/h)")
```

Subsampling and density contour plotting did not reveal separable populations. This graph clearly shows that female patients tended to have higher ESR levels than male patients, and that older patients tended to have higher ESR levels than younger patients. However, this effect was not universal, and many old patients of both genders had low ESR values. 


Do the HUP reference ranges appear to appropriately capture patients with normal ESRs? Abnormal ESRs?

```{r}
esr_sample <- esr_sample %>%
  mutate(
    HUP = case_when(
      (SEX == "Male" & RESULT > 15) | 
        (SEX == "Female" & AGE <= 50 & RESULT > 25) |
        (SEX == "Female" & AGE >50 & RESULT > 30) ~ "high",
      TRUE ~ "normal") %>% 
      factor(levels = c("high", "normal"))
  )

ggplot(data = esr_sample, 
       mapping = aes(x = AGE, 
                     y = RESULT, 
                     color = HUP)) +
  geom_point(size = 0.5, 
             alpha = 0.2) +
  geom_density2d(color = "black", 
                 size = 0.3) +
  facet_wrap(~SEX) + 
  labs(x = "Patient Age (years)",
       y = "ESR (mm/h)") + 
  guides(color = FALSE)
```

HUP's normal reference range appears to capture the majority of normal ESR results. It does not capture the tendency of older individuals to have higher ESR values. However, this tendency is not universal, and it is uncertain whether high ESR values in older individuals are physiologically normal.

#### ARUP Laboratories

```{r}
esr_sample <- esr_sample %>%
  mutate(
    ARUP = case_when(
      SEX == "Male" & RESULT >10 ~ "high",
      SEX == "Female" & RESULT >20 ~ "high",
      TRUE ~ "normal") %>% 
      factor(levels = c("high", "normal"))
  )

ggplot(data = esr_sample, 
       mapping = aes(x = AGE, 
                     y = RESULT, 
                     color = ARUP)) +
  geom_point(size = 0.5, 
             alpha = 0.2) +
  geom_density2d(color = "black", 
                 size = 0.3) +
  facet_wrap(~SEX) + 
  labs(x = "Patient Age (years)",
       y = "ESR (mm/h)") + 
  guides(color = FALSE) + 
  ggtitle("ARUP Laboratories")
```

#### Quest Diagnostics

```{r}
esr_sample <- esr_sample %>%
  mutate(
    Quest = case_when(
      (SEX == "Male" & AGE <=50 & RESULT >15) |
        (SEX == "Male" & AGE >50 & RESULT >20) |
        (SEX == "Female" & AGE <=50 & RESULT >20) |
        (SEX == "Female" & AGE >50 & RESULT >30) ~ "high",
      TRUE ~ "normal") %>% 
      factor(levels = c("high", "normal"))
  )

ggplot(data = esr_sample, 
       mapping = aes(x = AGE, 
                     y = RESULT, 
                     color = Quest)) +
  geom_point(size = 0.5, 
             alpha = 0.2) +
  geom_density2d(color = "black", 
                 size = 0.3) +
  facet_wrap(~SEX) + 
  labs(x = "Patient Age (years)",
       y = "ESR (mm/h)") + 
  guides(color = FALSE) + 
  ggtitle("Quest Diagnostics")
```

#### LabCorp

```{r}
esr_sample <- esr_sample %>%
  mutate(
    LabCorp = case_when(
      (SEX == "Male" & AGE <=50 & RESULT >15) |
        (SEX == "Male" & AGE >50 & RESULT >30) |
        (SEX == "Female" & AGE <=50 & RESULT >32) |
        (SEX == "Female" & AGE >50 & RESULT >40) ~ "high",
      TRUE ~ "normal") %>% 
      factor(levels = c("high", "normal"))
  )

ggplot(data = esr_sample, 
       mapping = aes(x = AGE, 
                     y = RESULT, 
                     color = LabCorp)) +
  geom_point(size = 0.5, 
             alpha = 0.2) +
  geom_density2d(color = "black", 
                 size = 0.3) +
  facet_wrap(~SEX) + 
  labs(x = "Patient Age (years)",
       y = "ESR (mm/h)") + 
  guides(color = FALSE) + 
  ggtitle("LabCorp")
```

#### Bakerman's ABCs

```{r}
esr_sample <- esr_sample %>%
  mutate(
    Bakerman = case_when(
      (AGE <18 & RESULT >13) ~ "high",
      (SEX == "Male" & AGE >=18 & AGE <40 & RESULT >15) |
        (SEX == "Female" & AGE >=18 & AGE <40 & RESULT >20) |
        (SEX == "Male" & AGE >=40 & RESULT > (AGE/2)) |
        (SEX == "Female" & AGE >=40 & RESULT > (AGE/2 + 5)) ~ "high",
      TRUE ~ "normal") %>% 
      factor(levels = c("high", "normal"))
  )

ggplot(data = esr_sample, 
       mapping = aes(x = AGE, 
                     y = RESULT, 
                     color = Bakerman)) +
  geom_point(size = 0.5, 
             alpha = 0.2) +
  geom_density2d(color = "black", 
                 size = 0.3) +
  facet_wrap(~SEX) + 
  labs(x = "Patient Age (years)",
       y = "ESR (mm/h)") + 
  guides(color = FALSE) + 
  ggtitle("Bakerman's ABCs")
```














