---
title: "Reproducible Report Template"
output: html_notebook
--- 

```{r setup, include=FALSE}
library(tidyverse)
library(odbc)
library(lubridate)
```

## Background and Objectives

This section should address the following questions:

* What is the question to be addressed by the project?
* How could addressing this question change patient management or further generalizable knowledge?
* What data resources could be used?
* What are the first 1-2 key graphs or statistics that should be obtained?


## Data Acquisition

* For database sources, include the following:
    * Database server address
    * Database table name
    * No login credentials!
* For data file sources, include the following:
    * Origin, with references
    * File name and format
* For all data sources, include the following:
    * General properties of the data set
    * Code Book

### Sample Code

#### Load CSV file

```{r}
my_data <- read_csv("my_data.csv")
```

#### Load Excel file

```{r}
library(readxl)
my_data <- read_excel("my_data.xlsx")
```

#### Connect to the MGH Lab Utilization Datamart

```{r connect-datamart}
username <- function() {
    .rs.api.showPrompt(title = "Username", 
                       message = "Please enter your Partners User Name")
}

password <- function() {
    .rs.api.askForPassword(prompt = "Please enter your Partners Password")
}

con <- dbConnect(odbc(),
                 dsn = "PHSSQL2057",
                 uid = str_c("PARTNERS\\", username()),
                 pwd = password())
  
labresults_tbl <- tbl(con, "MGHLABUTIL_LabResults")
```

#### List all test order names, ordered by frequency

```{r}
tests <- labresults_tbl %>%
    select(TstOrderName, CollectDateTime) %>%
    filter(year(CollectDateTime) == "2017") %>%
    group_by(TstOrderName) %>%
    summarize(count = n()) %>%
    arrange(desc(count)) %>%
    collect()
```

#### Database Query

```{r query}
esr <- labresults_tbl %>%
  select(column1,
         column2,
         ...) %>%
  filter(condition1,
         condition2,
         ...,
         year(CollectDateTime) == "2017") %>%
  arrange(CollectDateTime) %>%
  collect()
```

#### Data Dump

```{r save-rds}
esr %>% 
    write_rds(str_c("my_data_", today(), ".rds"))
```

#### Disconnect

```{r}
dbDisconnect(con)
```


## Data Exploration and Clean-Up

* Convert messy data to tidy
    * Each variable in its own column
    * Each case in its own row
    * Each value in its own cell
* Check for missing values
* Then, go column by column and do the following:
    * Verify (and if necessary, convert) data type
        * Numeric variables should be integer or numeric
        * Date-time variables should be Date or POSIXct
        * Categorial variables should be factor
        * All others should be character
    * Visualize distribution
        * Numeric variables: histogram
        * Categorical variables: bar graph
        * Character variables: inspect a small sample
    * Comment on whether the data is as expected
        * Range and shape of distribution
        * Number and breakdown of categories
        * Properties of character variables
    * Explain and/or fix discrepancies

### Sample Code

#### Missing values

```{r}
esr %>%
    map_df(function(x) sum(is.na(x))) %>%
    gather(column, NAs)
```

#### Numeric variables

`variable_1` is an **integer** column, which is appropriate.

```{r}
ggplot(data = my_data) + 
    geom_histogram(mapping = aes(x = variable_1),
                   binwidth = 1)
```

#### Factor variables

`variable_2` is a **factor** column, which is appropriate.

```{r}
ggplot(data = my_data) + 
    geom_bar(mapping = aes(x = variable_2))
```

#### Character variables

`variable_3` is a **character** column, which is approprite.

```{r}
set.seed(1)
my_data %>%
    sample_n(10) %>%
    pull(variable_3)
```

#### Convert from character to factor

`variable_4` is a **character** column, but since it represents a categorial variable, convert it to a **factor**.

```{r}
my_data <- my_data %>%
    mutate(variable_4 = as_factor(variable_4))
```

#### Convert from character to integer

`variable_5` is a **character** column, but since it represents a numeric variable, convert it to **integer**.

Before doing so, examine any rows that do *not* contain an integer value:

```{r}
my_data %>%
    filter(is.na(as.integer(variable_5))) %>%
    group_by(variable_5) %>%
    summarize(n = n()) %>% 
    arrange(desc(n))
```

Now convert to integer and drop all rows in which `variable_5` had a non-numeric entry.

```{r}
my_data <- my_data %>%
    mutate(variable_5 = as.integer(variable_5)) %>%
    drop_na(variable_5)
```

## Visualization and Modeling


## References

