---
title: "Course Project Template"
output: html_notebook
--- 

The purpose of this template is:

1. to recap some of the concepts covered in the course, and 
2. to provide a resource of code snippets that will be useful for your course project (and hopefully other future R projects!)

As you prepare the report, you will likely encounter strange errors and get stuck multiple times. **That's normal and there's no way around it as you learn**. The basic approach for getting un-stuck is to spend a **limited** amount of time **actively** looking for the solution and then **asking for help**. 

Two common scenarios, and a systematic way to troubleshoot them are:

1. **You encounter an unexpected result or error**
    * Check **spelling** - missing or extra quotes, commas, parentheses, spaces, and misspelled function names
    * Read the error message and try to make sense of it
    * Google the error message (cut and paste!)
2. **You (kind of) know what you want to do but don't know how to do it in R**
    * Consult a relevant [RStudio Cheat Sheet](https://www.rstudio.com/resources/cheatsheets/)
    * If you know it was covered in the workshop, re-watch the workshop. The shared drive for Penn Pathology is: smb://uphsfp15.uphs.upenn.edu/SHAREDATA2/HUP/Pathology/Reproducible_Clinical_Data_Analysis
    * Do a Datacamp tutorial. Recommended tutorials include:
        * [Introduction to the Tidyverse](https://www.datacamp.com/courses/introduction-to-the-tidyverse) is highly recommended for everyone. 
        * [Data Visualization with ggplot2 (Part 1)](https://www.datacamp.com/courses/data-visualization-with-ggplot2-1) for more practice with ggplot.
        * [Data Manipulation in R with dplyr](https://www.datacamp.com/courses/dplyr-data-manipulation-r-tutorial) for more practice with dplyr
        * [Cleaning Data in R](https://www.datacamp.com/courses/cleaning-data-in-r) is useful if you have messy data that needs to be cleaned.
        * [Joining Data in R with dplyr](https://www.datacamp.com/courses/joining-data-in-r-with-dplyr) is useful if you need to combine data from two separate data frames.

Asking good questions when you get stuck is a key skill for coding and collaborative data analysis. I recommend the following formula:

>  I tried to accomplish (*goal*) so I tried this code: 
>  (*copy and paste the code here*)
>  
>  I expected this: ...
>  
>  Instead I get this result / error: 
>  (*copy and paste the result or error message*)

The best resources for asking questions are:

1. Email one of the course directors: [Stephan](mailto:kadaukes@email.chop.edu), [Amrom](mailto:obstfelda@email.chop.edu), or [Dan](mailto:Daniel.Herman2@uphs.upenn.edu)
2. Post your question on the [CHOP R User Group Slack channel](https://choprusergroup.slack.com)

What follows is a suggested structure for a reproducible report as well as some code snippets that you might want to use or modify.


## Background and Objectives

This section should address the following questions:

* What is the question to be addressed by this project?
* How could addressing this question change patient management and/or further generalizable knowledge?
  * What's the point of doing this analysis? 
  * Would anyone else care about the results?
* What are the first 1-2 simple but important graphs or statistics that you'd like to generate?


## Data Acquisition

The **narrative** of this section should cover the following:

* General properties of the data set
* For data files:
    * Origin, with details on how is was created and/or source from where it was acquired
    * File name and file format
* For databases:
    * Database server, database, and table name (**not** login credentials!)
* Code Book
    * List each column of the data set (after isolating data of interest) with a concise description. 

The **code** in this section should accomplish the following:

* For data files:
    * Import your data into R
    * Isolate the data of interest
* For databases:
    * Connect to the database
    * Query the data and place it into a data frame
    * Optional: Dump the data in a local file
    * Disconnect

### Import data file

Don't forget to load necessary packages first:

```{r}
library(tidyverse)
library(lubridate)
```

#### CSV file

Here is a sample code snippet to load a CSV file named `my_data.csv` into a data frame named `my_data`. Try to give your data more descriptive names than that!

```{r}
my_data <- read_csv("my_data.csv")
```

#### Excel file

This code snippet loads and Excel file `my_data.xlsx` into a data frame named `my_data`. We need to load the `readxl` package for the `read_excel()` function to work.  

```{r}
library(readxl)
my_data <- read_excel("my_data.xlsx")
```

### Isolate data of interest

This sample code snippet would select the `Patient_MRN`, `Patient_Age`, `Patient_Sex`, `Date_of_Collection`, and `Result` columns and then filter the rows such that only those rows remain in which `Date_of_Collection` is within the year of 2017. The `arrange()` function would then sort rows by `Date_of_Collection`. 

```{r}
my_data <- my_data %>%
  select(Patient_MRN,
         Patient_Age,
         Patient_Sex,
         Date_of_Collection,
         Result) %>%
  filter(year(Date_of_Collection) == 2017) %>%
  arrange(Date_of_Collection)
```

### Code book

The code book is where you document concisely document the data contained in each column. Be sure to include enough detail to avoid misunderstandings. For example:

- **Patient_MRN**. Patient medical record number (HUP)
- **Patient_Age**. Patient age (in years) at the time the specimen was collected.
- **Patient_Sex**. Patient sex. M = Male, F = Female.
- **Date_of_Collection**. The date on which the specimen was collected.
- **Result**. Laboratory result value for this specimen.

## Data Exploration and Cleaning

The **narrative and code** of this section should, together, describe and accomplish the following:

1. Convert **messy** data to **tidy**
2. Check for **missing values**
3. **Explore** and **fix** data in each column
4. **Display** the cleaned data table

### Convert messy data to tidy

If any of following are not true for your data set, then you will have to tidy your data:

* Each **variable** is in its own **column**
* Each **case** is in its own **row**
* Each **value** is in its own **cell**

A tutorial for cleaning data can be found here: [Cleaning Data in R](https://www.datacamp.com/courses/cleaning-data-in-r)

### Missing values

The following code chunk shows columns that have missing data in the data frame `my_data`. 

```{r}
my_data %>%
    map_df(function(x) sum(is.na(x))) %>%
    gather(column, NAs)
```

### Explore and fix data column-by-column

In this section, go column by column and do the following:

* Verify (and if necessary, convert) data type
    * **Numeric** variables should be **integer** (whole numbers) or **numeric** (fractions)
    * **Date-time** variables should be **Date** (whole dates) or **POSIXct** (date/time stamps)
    * **Categorial** variables should be **factor**
    * All others, in general, should be **character**
* Examine the distribution
    * **Numeric** variables: **histogram**
    * **Categorical** variables: **numerical summary**
    * **Character** variables: inspect a small **sample**
* Comment on whether the data is as expected
    * Range and shape of distribution
    * Number and breakdown of categories
    * Properties of character variables
* Explain discrepancies
* Resolve data problems

Below are a few examples of how various types of columns could be handled.

#### A numeric column

`Patient_Age` is an **integer** column, which is appropriate.

```{r}
ggplot(data = my_data) + 
    geom_histogram(mapping = aes(x = Patient_Age),
                   binwidth = 1)
```

The histogram shows ..., which is expected.

#### A character column

`Patient_MRN` is a **character** column, which is appropriate.

```{r}
set.seed(1)
my_data %>%
    sample_n(10) %>%
    pull(Patient_MRN)
```

A sample of data shows ..., which is expected.

#### A character column that should be a factor

`Patient_Sex` is a **character** column, but since it represents a categorial variable, convert it to a **factor**.

```{r}
my_data <- my_data %>%
    mutate(Patient_Sex = as_factor(Patient_Sex))
```

```{r}
my_data %>%
  pull(Patient_Sex) %>%
  summary()
```

A numerical summary shows ..., which is expected.

#### A character column that should be integer

`Result` is a **character** column, but since it represents a numeric variable, convert it to **integer**.

Before doing so, examine any rows that do **not** contain an integer value:

```{r}
my_data %>%
    filter(is.na(as.integer(Result))) %>%
    group_by(Result) %>%
    summarize(n = n()) %>% 
    arrange(desc(n))
```

Now convert to integer and drop all rows in which `Result` had a non-numeric entry.

```{r}
my_data <- my_data %>%
    mutate(Result = as.integer(Result)) %>%
    drop_na(Result)
```

```{r}
ggplot(data = my_data) + 
    geom_histogram(mapping = aes(x = Result),
                   binwidth = 1)
```

The histogram shows ..., which is expected.

### Overview of the cleaned data

The following code chunk creates an interactive table from the `my_data` data frame, which is useful for peer review and presentation purposes.

```{r}
my_data
```

## Visualization and Modeling

This section should iterate through multiple cycles of: 

1. Pose a **question** or state a **hypothesis**
2. Create a **graph** or **statistical summary** to address #1
3. **Comment** on new insights gained and new problems raised


## Summary and Conclusions

