---
title: "Getting Data"
output: html_notebook
---

## Your Turn #2: Importing a CSV file

We have placed a file on the RStudio server that contains lab data for a specific test (ESR) generated here at Penn. The file is named "esr.csv" and you should be able to see it in your *Files* pane.

Inside the code chunk below, add a line of code that accomplishes the following:

  1. Loads the CSV file "esr.csv"
  2. Stores the resulting data frame in an object named "esr" (without the quotes)

Hint: this is very similar to what's on the slide titled "read_csv()"!

```{r}
library("tidyverse")

```

Remember that in order to run a code chunk, you need to click on the little green triangle on the upper right corner of the chunk.

Note: When you run the code chunk, you might see some R output that at first glance might look like an error, with red type stating "Parsed with column specification: ..." This is actually just a message from read_csv() telling you what it guessed the data types in the indicated columns are, e.g. AGE = col_double (numeric) or DRAWN_DT_TM = col_datetime (date-time stamp). 


### Viewing a data frame in RStudio

One thing that Excel does well is to provide an interactive visual representation of the data, allowing you inspect it by sorting and filtering. RStudio actually does this well, too, with one difference - it won't let you *change* any of the data while you inspect it.

Look on the right at the *Environment* pane (you might have to click on the "Environment" tab) and find the entry *esr*. This is the data frame you just created inside of R's memory. 

On the far right in this row, you will spot a symbol that looks like a small table. Click on it to make the Table Viewer appear.

*Pair up* with your group mates and answer the following questions:

  1. How many rows are in the data frame? How many columns?
  
  2. What does each row represent? Each column?

  3. Look at the two columns labeled *RESULT_VALUE_NUMERIC* and *ASCII_TEXT*. Can you figure out what the difference is between the two? 

  4. Go ahead and try to edit one of the values in this viewer. You will find that you can't. It would have been easy for the RStudio programmers to allow editing of specific values, but they decided not to add that feature. Why do you think this was designed that way?

  5. Can you think if an easy way to confirm that all samples were collected in between *05-18-2008* and *08-14-2018*?

  6. Can you figure out, *without writing any R code*, how many of the rows come from *female* patients?


*Stop here.* We will continue later.

***************



## Your Turn #4: Databases

### Browsing a Database

To connect to the database containing ESR values, find the *Connections* tab on the top right pane and click *New Connection*. From there, select *LABS* and click OK. You have now connected to the a *Database Management System* with real patient lab data that we've set up for this course. (So don't worry, you're not going to be able to break anything important!)

You should now see two *Databases*, one of which is titled *public*. Click on the blue circle with the white triangle to see the *Database Tables*. Only one table should appear - *RESULTS*.

Click on the blue circle with the white triangle next to *RESULTS*. This reveals all the columns (or variables) in the table. Some of the column names are obvious, such as *HUP_MRN* while others are more cryptic. 

On the far right next to *RESULTS*, you will spot a symbol that looks like a small table, just like the one you saw ealier next to the *esr* data frame. Click on it to *preview* the database table. This will query a number of rows from the database server and display it in RStudio.

*Pair up* with your group mates and answer the following questions:

  1. How many rows did RStudio retrieve from the database? Why do you think RStudio only pulled a small subset of all the the rows in *RESULTS*?

  2. Suppose you want to create a graph that shows the results of all ESR values from 2017 versus patient age and patient sex. The graph should be a scatterplot in which each data point is a specific lab result. Without looking at the specific column names in *RESULTS*, what information do you think you will need?

  3. Try to identify the specific columns in *RESULTS* that contain this information from question 2.

  4. Even though the patient name and MRN are not going to be necessary for creating the graph, why might it be a good idea to pull this information as well?

  5. Once you are done browsing, please *disconnect* by clicking the button that looks like a pipe with a little red X on it.
  


### Connecting to a Database with Code

The most versatile method to connect to databases with R involves the `odbc` package. ODBC stands for "Open Database Connectivity" and is a widely used standard to connect to databases. Almost every database management system comes with an ODBC driver, and therefore R can connect to almost every type of database.

Before you can import data from a database table into R with `odbc`, you need to complete two steps:

1. Create a *database connection object* that links to the database
2. Create a *tbl object* that links to a specific database table

The tbl object can be thought of as a virtual dataframe - it can be used in lieu of an actual dataframe that exists in R's memory, but it's really just a reference to the database table. When you print out values from a tbl object, as we will do below, what happens inside of R is that R sends a database query to the database to obtain the desired data.


The following code chunk opens a database connection to the *LABS* database and creates a database connection object named `labs_con`. 

```{r}
library(odbc)
labs_con <- dbConnect(odbc(), "LABS")
```



After running this code chunk, you will notice that a new object shows up in the Environment, named `labs_con`, with the data type "Formal class PostgreSQL". If you click on the blue triangle to expand it, you will see information about the connection. For example, that the username of the connection is "postgres" and the server name "localhost". 

How does R know these details about the *LABS* database? The database connection details are stored in a configuraton file called "odbc.ini" in a folder called "/etc" (the standard Linux folder for system configuration files). Try to find the file and open it. If you get stuck, ask for help!

Take a look at the contents of the "odbc.ini" file. *Pair up and discuss* - do the contents make sense to you? Why might it be a good idea to store the connection details in a separate file from your R Markdown document?


Now that our database connection is set up, the next code chunk creates a tbl object named `results_tbl` that links to the *RESULTS* database table.

```{r}
results_tbl <- tbl(labs_con, "RESULTS")
```


We can look at the first 1,000 rows of the database table by simply writing:

```{r}
results_tbl
```


Please disconnect from the database when you're done. (This has the same effect as clicking the button in the Connections pane that looks like a pipe with a little red X on it.)

```{r}
dbDisconnect(labs_con)
```