---
title: "Exploring Data"
output: html_notebook
---

## Your Turn #2 - Visualizing Data

### Import ESR Data

Run the following code to import the practice data set.

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(lubridate)
esr <- read_csv("esr.csv") %>%
  filter(year(DRAWN_DT_TM) == 2017)
```

### Practice ggplotting

Recall the histogram of ESR values from the beginning of the session. You will now create it yourself! You will need to:

1. Pick a data set: *esr*
2. Choose a geom function: *geom_histogram()*
3. Map aesthetic properties of the graph to variables of the data frame (in this case, the "RESULT_VALUE_NUMERIC" variable should be on the x axis; on histograms the y axis always represents the number of values that fall within an interval): *x = RESULT_VALUE_NUMERIC*

Fill in the blanks and run the code to create the plot.

(Hint: if you get stuck, take a look at the slide titled "ggplot()"!)

```{r, message = FALSE}
ggplot(data = ______) +
  ______(mapping = aes(______)) + 
  xlim(0, 150)
```

Congratulations, you created your first ggplot!




Make a scatter plot with "AGE" on the x axis and "RESULT_VALUE_NUMERIC" on the y axis. (Hint: Look at the Cheat Sheet to figure out which geom_ function to use to make a scatter plot!)

```{r, message = FALSE, warning = FALSE}
ggplot(data = ______) +
  ______(mapping = aes(x = ______, y = ______)) +
  ylim(0,150)
```





Can you figure out how to modify the above code to draw a contour plot (aka. 2D density plot)? Use the cheat sheet!

```{r}

```





To the above plot, add a third aesthetic, *color*, to indicate patient sex (SEX).

```{r}

```


Stop here. 


******************



## Your Turn #3: Isolating Data

### select()

Fill in the blanks below to have R select the "AGE" and "SEX" columns from the "esr" data frame. 

```{r}
select(______, ______, ______)
```




Select the "HUP_MRN", "PT_NAME", and "RESULT_VALUE_NUMERIC" columns from "esr".

```{r}

```




### filter()

Show all rows where the variable "AGE" is greater than or equal to 100.

```{r}
filter(esr, __________)
```




Show all rows that have a missing value for "RESULT_VALUE_NUMERIC". (Hint: Look at the slide titled "Logical Tests"!)

```{r}
______(______, ______)
```




Rewrite the above code so that it uses the pipe (%>%) operator to plug "esr" into the filter() function. (Hint: Look at the slide titled "The Pipe Operator %>%"!)

```{r}

```




#### Combined logical tests

There are three operators which allow you to combine two or more logical tests.

   &  "and"
   |  "or"
   !  "not"

For example, the following code uses the "and" (&) operator to select rows in which "SEX" is neither "Female" nor "Male".

```{r}
filter(esr, SEX != "Female" & SEX != "Male")
```




Show all rows where "SEX" is "Male" *and* "RESULT_VALUE_NUMERIC" is greater than 13.

```{r}
filter(esr, ______)
```




Rewrite the above code so that it uses the pipe (%>%) operator to plug "esr" into the filter() function.

```{r}

```




### arrange()

Fill in the blank below to sort by "DRAWN_DT_TM".

```{r}
arrange(esr, ______)
```




Now change the order so that most recently drawn samples are shown first.

```{r}
arrange(______, ______)
```




Rewrite the above code so it uses the pipe operator to plug "esr" into arrange().

```{r}

```




### Pipelines

Now let's try to construct a pipeline that give us an ordered list of results from women with a normal ESR value, by age.

Write a pipeline that:

1. Starts with the data frame "esr", then
2. Selects the "AGE", "SEX", and "RESULT_VALUE_NUMERIC" columns, then
3. Filters so only rows in which "SEX" is equal to "Female" remain, then
4. Filters so only rows in which "AGE" is smaller than or equal to 20 remain, then
5. Arranges the results first by age (in ascending order), then as a tie-breaker by ESR result (in descending order).

```{r}

```

Stop here.


******************

## Your Turn #4: Grouping and Summarizing Data 

In this exercise, we will find the number of distinct patients in the the "esr" data set and will find the average and standard deviation of the age of patients, broken down by patient sex.

One problem is that many patients in "esr" have multiple results. We don't want the results to be skewed towards patients who have multiple results logged. To solve this problem, we will take advantage of the dplyr function "distinct", which removes duplicate values in a data frame.

Can you figure out how to use distinct() to answer the following: How many distinct patients are represented in "esr"?

```{r}

```




Next we will use the set of distinct patients and pipe them into the group_by() and summarize() functions to compute three summary statistics about "esr", separately for men and women (i.e. grouped by "SEX"):

1. The number of distinct patients.
2. The average (mean) age of patients.
3. The standard deviation of the age of patients.

Note: we need to pass the option ".keep_all = TRUE" to distinct() because otherwise it will drop all the columns except for the ones for which we select unique values.

```{r}
esr %>%
    distinct(______, .keep_all = TRUE) %>%
    group_by(______) %>%
    summarize(MeanAge = ______,
              SDAge   = ______,
              nPts    = ______)
```

If you've gotten this code chunk to work, great job! Do the results surprise you? Why or why not? 
